---
title: "UMAP projection of globe"
author: "Lambda Moses"
date: "1/20/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, eval=FALSE}
# Install the packages. Everything is on CRAN, except that bluster is on Bioconductor
install.packages(c("uwot", "SphericalK", "tidyverse", "sf", "plotly", "gganimate", 
                   "spData", "igraph", "BiocManager"))
BiocManager::install("bluster")
```

```{r}
library(uwot)
library(SphericalK)
library(tidyverse)
library(sf)
library(plotly)
library(gganimate)
library(spData)
library(igraph)
library(bluster)
```

# Generate the random points on sphere
The points should be uniformly distributed on the sphere, so using `runif` on the longitudes and latitudes will not work because in that case, there will be higher density of points around the poles. But I can use the `SphericalK` package to generate random points uniformly distributed on a sphere.
```{r}
points <- sphere_random(10000)
points <- as.data.frame(points)
head(points)
```
```{r}
ggplot(points, aes(longitudes, latitudes)) +
  geom_point(size = 0.5, alpha = 0.5) +
  coord_equal()
```

I can't color 200 countries and keep the colors discernible, so I'll color by continent.
```{r}
# Colorboind friendly palette from dittoSeq
pal <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", 
         "#CC79A7", "#666666", "#AD7700", "#1C91D4", "#007756", "#D5C711", 
         "#005685", "#A04700", "#B14380", "#4D4D4D", "#FFBE2D", "#80C7EF", 
         "#00F6B3", "#F4EB71", "#06A5FF", "#FF8320", "#D99BBD", "#8C8C8C", 
         "#FFCB57", "#9AD2F2", "#2CFFC6", "#F6EF8E", "#38B7FF", "#FF9B4D", 
         "#E0AFCA", "#A3A3A3", "#8A5F00", "#1674A9", "#005F45", "#AA9F0D", 
         "#00446B", "#803800", "#8D3666", "#3D3D3D")
pal_use <- pal[c(1:7, 9, 8)]
```

```{r}
ggplot(world) +
  geom_sf(aes(fill = continent)) +
  scale_fill_manual(values = pal_use)
```

```{r}
st_crs(world)
```

```{r}
points_sf <- points %>% 
  st_as_sf(coords = c("longitudes", "latitudes")) %>% 
  st_set_crs("EPSG:4326")
```

Check that the polygons for countries are valid as that's required in `st_join`.
```{r}
all(st_is_valid(world))
```

So the polygons are all valid. Get the continents
```{r}
points_sf <- points_sf %>% 
  st_join(world[, c("continent", "geom")]) %>% 
  mutate(continent = case_when(is.na(continent) ~ "oceans",
                               TRUE ~ continent),
         continent = fct_relevel(continent, "oceans", after = Inf))
world <- world %>% 
  mutate(continent = fct_expand(continent, "oceans"),
         continent = fct_relevel(continent, "oceans", after = Inf))
```

```{r}
ggplot() +
  geom_sf(data = world, aes(color = continent)) +
  geom_sf(data = points_sf, aes(color = continent), size = 0.5) +
  scale_color_manual(values = pal_use, drop = FALSE)
```

# Get 3D coordinates
I know, the Earth is not exactly a sphere, and there's such a thing called datum for how the Earth is not exactly a sphere. But for the purpose of this animation, I'll just use a sphere. Say the radius is 6371 km, according to this: https://nssdc.gsfc.nasa.gov/planetary/factsheet/earthfact.html
```{r}
points_sf <- cbind(points_sf, st_coordinates(points_sf))
```

```{r}
r <- 6371
points_sf <- points_sf %>% 
  rename(lat = Y, lon = X) %>% 
  mutate(lat = lat*pi/180, # Convert to radians
         lon = lon*pi/180,
         x = r * cos(lat) * cos(lon),
         y = r * cos(lat) * sin(lon),
         z = r * sin(lat))
```

```{r}
plot_ly(points_sf, x = ~x, y = ~y, z = ~z, color = ~continent,
        colors = pal_use) %>% 
  add_markers(marker=list(size = 3))
```

Cool, that's what I want.

# UMAP projection
See how it looks with default parameters
```{r}
set.seed(29)
world_umap <- umap(st_drop_geometry(points_sf[,c("x", "y", "z")]))
colnames(world_umap) <- c("UMAP1", "UMAP2")
```

```{r}
umap_df <- cbind(as.data.frame(world_umap), 
                 st_drop_geometry(points_sf[,"continent", drop = FALSE]))
```

```{r}
ggplot(umap_df, aes(UMAP1, UMAP2, color = continent)) +
  geom_point(size = 0.5) +
  scale_color_manual(values = pal_use) +
  coord_equal() +
  guides(color = guide_legend(override.aes = list(size = 2)))
```

The points are really clumped, mostly by continent. Now I'll do what was done here: https://jef.works/blog/2022/01/19/exploring-umap-parameters/
I made the code from the blog more concise and relevant to the case here.
```{r}
run_umap_param_grid <- function(X, umap_params) {
  map_dfr(seq_len(nrow(umap_params)), function(i) {
    print(i)
    emb <- uwot::umap(
      X         = X,
      spread    = umap_params$spread[i],
      min_dist  = umap_params$min_dist[i],
    )
    colnames(emb) <- c("UMAP1", "UMAP2")
    emb <- as.data.frame(emb)
    emb$spread <- umap_params$spread[i]
    emb$min_dist <- umap_params$min_dist[i]
    emb$param_set <- paste0("min_dist: ", umap_params$min_dist[i], 
                            ", spread: ", umap_params$spread[i])
    emb
  })
}
```

```{r}
plot_umap_param_grid <- function(umaps, color_name) {
  ggplot(umaps) +
    geom_point(aes_string(x = "UMAP1", y = "UMAP2", color = color_name), 
               size = 0.5) +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank()
    ) +
    scale_color_manual(values = pal_use) +
    facet_wrap(min_dist ~ spread , 
               labeller = label_both,
               scales = "free") +
    guides(color = guide_legend(override.aes = list(size = 2))) +
    theme_minimal()
}
```

```{r}
## try some different parameters
umap_params <- expand.grid(
	spread   = c(0.5, 1, 4),
	min_dist = c(0.01, 0.15, 0.75)
)
umaps <- run_umap_param_grid(st_drop_geometry(points_sf[,c("x", "y", "z")]),
                             umap_params)
umaps$continent <- rep(points_sf$continent, 9)
plot_umap_param_grid(umaps, "continent")
```

Interesting. With larger min_dist, and with the oceans, the projection kind of looks like the map.

Why shall I not animate the transition through all 9 sets?
```{r}
anim_umap_param_grid <- function(umaps, color_name = NULL) {
  p <- ggplot(umaps, aes(UMAP1, UMAP2)) +
    coord_equal()
  if (!is.null(color_name)) {
    p <- p +
      geom_point(aes_string(color = color_name), size = 0.5, alpha = 0.8) +
      scale_color_manual(values = pal_use) +
      guides(color = guide_legend(override.aes = list(size = 2)))
  }
  anim <- p +
    transition_states(param_set,
                      transition_length = 3,
                      state_length = 1) +
    labs(title = '{closest_state}') +
    theme(plot.title = element_text(size = 28)) +
    enter_fade()			
  anim + theme_minimal() +
    view_follow()
}
```

```{r}
anim_umap_param_grid(umaps, "continent")
```

```{r}
anim_save("world_umap.gif")
```

How about the animation but without coloring by continent? Then it's just UMAP projections of points uniformly distributed on a sphere.

```{r}
anim_umap_param_grid(umaps, NULL)
```

```{r}
anim_save("sphere_umap.gif")
```

How about color by Leiden clustering? I really don't expect meaningful clusters since the points are uniformly distributed on the sphere.
```{r}
g <- makeKNNGraph(st_drop_geometry(points_sf[,c("x", "y", "z")]), k = 20)
points_sf$leiden <- cluster_leiden(g, objective_function = "modularity",
                                   resolution_parameter = 0.2)$membership
umaps$leiden <- rep(points_sf$leiden, 9)
umaps$leiden <- factor(umaps$leiden, 
                       levels = as.character(1:length(unique(points_sf$leiden))))
```

```{r}
anim_umap_param_grid(umaps, "leiden")
```

```{r}
anim_save("leiden_umap.gif")
```

Where are the leiden clusters on the map?
```{r}
ggplot() +
  geom_sf(data = points_sf, aes(color = factor(leiden)), size = 0.5, alpha = 0.5) +
  geom_sf(data = world, fill = NA) +
  scale_color_manual(values = pal_use) +
  guides(color = guide_legend(override.aes = list(size = 2)))
```

# Without the ocean
```{r}
# Version without ocean, to better show Leiden clustering
points_sf_land <- points_sf %>% 
  filter(continent != "oceans") 
```

```{r}
umaps_land <- run_umap_param_grid(st_drop_geometry(points_sf_land[,c("x","y","z")]),
                                  umap_params)
```

```{r}
umaps_land$continent <- rep(points_sf_land$continent, 9)
```

```{r}
plot_umap_param_grid(umaps_land, "continent")
```

```{r}
anim_umap_param_grid(umaps_land, "continent")
```

```{r}
anim_save("world_umap_land.gif")
```

How about leiden clustering?
```{r}
g <- makeKNNGraph(st_drop_geometry(points_sf_land[,c("x", "y", "z")]), k = 20)
points_sf_land$leiden <- cluster_leiden(g, objective_function = "modularity",
                                        resolution_parameter = 0.2)$membership
points_sf_land$leiden <- factor(points_sf_land$leiden, 
                                levels = as.character(1:length(unique(points_sf$leiden))))
umaps_land$leiden <- rep(points_sf_land$leiden, 9)
```

```{r}
plot_umap_param_grid(umaps_land, "leiden")
```

How do the leiden clusters correspond to the continents? Without the ocean, I do expect meaningful clusters.
```{r}
ggplot() +
  geom_sf(data = points_sf_land, aes(color = leiden), size = 0.5) +
  geom_sf(data = world, fill = NA) +
  scale_color_manual(values = pal_use) +
  guides(color = guide_legend(override.aes = list(size = 2)))
```

That's just one set of Leiden clustering parameters.
```{r}
anim_umap_param_grid(umaps_land, "leiden")
```

```{r}
anim_save("leiden_umap_land.gif")
```
